GPU-Beschleunigung der Material Point Method

### 2018-11-11 mscthesis.tex
push-forward,pull-back, eulerian+lagrangian, notations

\cite{MPM:GPU}
%look again for MPM/PIC gpu impl
%SPH, FEM methods
\section{Notation}
Einstein-Notation
\section{Basics}
\subsection{Continuum Mechanics}
\cite{MPM:COURSE}
\subsection{Constitutive Models}
\cite{MPM:DRUCKER} %Sand Plasticity
\cite{MPM:MULTI} %Sand+Water
\cite{MPM:INVERT} %Neo-Hookean
\cite{MPM:OLROYDB} %Viscoelastic Foam,Sponge
\cite{MPM:SHELLS} %Thin Shells
\cite{MPM:PHASE_CHANGE} %Heat-Model
\cite{MPM:OPTIMI_INTEGR} %Opimization Methods, BDF-Semi-Impl-Midpoint rule

The first aim will be to find the strain energy density $\psi(\mathbf{\epsilon})$ of the strain $\bm{\epsilon}$.

@@

where $\mathbf{\sigma}$ and $\bm{\epsilon}$ are second-order tensors with (3x3) = 9 elements. $C_{ijkl}$ is a fourth-order tensor with (3x3)x(3x3) = 81 elements.

@@

  \sigma _ { i j } = \sigma _ { i j } ( \mathbf{\epsilon} ) = \frac { \partial { \psi } } { \partial \epsilon _ { i j } },

@@

  = \mathbf{C}^{-1} \bm{\sigma}

@@

using engineering constants Young's modulus $E$ and Poisson ratio $\nu$. Inverting $\mathbf{C}^{-1}$ and switching to Lamé parameters $\lambda$ and $\mu$ results in the equation:

@@

  \boldsymbol{\epsilon} = \frac { 1 } { 2 } \left( \mathbf { F } + \mathbf { F } ^ { T } \right) - \mathbf { I }

@@

The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t. to $\mathbf{F}$: $\psi(\mathbf{R}_{0} \mathbf{F}) \neq \psi(\mathbf{F})$. The deformation caused by stress however does not affect rigid body motions. Given the polar decomposition $\mathbf{F} = \mathbf{RS}$, an alternate strain measure may be defined as:

@@

  \hat { \boldsymbol { \epsilon } } ( \mathbf { F } ) = \boldsymbol { \epsilon } \left( \mathbf { R } ^ { T } \mathbf { F } \right) = \frac { 1 } { 2 } \left( \mathbf { R } ^ { T } \mathbf { F } + \left( \mathbf { R } ^ { T } \mathbf { F } \right) ^ { T } \right) - \mathbf { I } = \mathbf { S } - \mathbf { I }

@@

Substituting $\hat{\boldsymbol{\epsilon}}$ into equation \ref{eq:linear_elas} leads to the energy definition of corotational hyperelasticity using rotation-invariance of Frobenius norm and trace:

@@

  \Psi = \mu \| \mathbf { F } - \mathbf { R } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \mathbf { R } ^ { T } \mathbf { F } - \mathbf { I } \right)

@@

\begin{equation}
  \Psi(\boldsymbol{\Sigma}) = \mu \| \boldsymbol { \Sigma } - \mathbf { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma } - \mathbf { I } )
\end{equation}
\cite{ADAMS:ELASTICITY}
\cite{MPM:INVERT}

\subsection{Singular Value Decomposition}
\cite{SVD:3x3}
\subsection{Material Point Method}
\subsubsection{Conservation Equations}
\cite{MPM:APIC}
\subsubsection{CFL Condition}
\subsection{GPU memory}
\subsubsection{Optimization opportunities}
\cite{NVIDIA:BEST:PRACTICE}
\cite{AMD:GPU_OPEN}
\begin{itemize}
  \item Minimize CPU-GPU transfers
  \item Coalesced memory access
  \item Bank conflicts
\end{itemize}
\subsubsection{SoA vs. AoS}
\section{Implementation}
\newcommand{\var}[1]{{\ttfamily#1}}% variable
\begin{algorithm}[t]
  \caption{Euclid’s algorithm}\label{euclid}
  \begin{algorithmic}[1]
    \Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
    \State $r\gets a\mathbfod b$ and \var{foobar}\label{foobar}
    \While{$r\not=0$}\Comment{We have the answer if r is 0}
    \State $a\gets b$
    \State $b\gets r$
    \State $r\gets a\mathbfod b$
    \EndWhile\label{euclidendwhile}
    \State \textbf{return} $b$\Comment{The gcd is b}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}
\subsection{Framework}
decorator pattern
\subsection{Parallel Reduction}
\cite{NVIDIA:PARALLEL_REDUCTION}
\subsection{Binning}
\cite{NVIDIA:BINNING}
\subsection{Scanning}
\cite{NVIDIA:SCAN}
\cite{NVIDIA:SCAN_MODERN}
\subsection{Affine particle-in-cell}
\cite{MPM:APIC}
\subsection{Iterative solvers for linear systems}
%GMRES, Conjugate Gradients
\subsection{Verifying Simulation}
\begin{itemize}
  \item Conservation equations
  \item Singular Value Decomposition
  \item Parallel Reduction
\end{itemize}
\section{Evaluation}
\cite{NVIDIA:PEAK_PERFORMANCE}
\cite{KHRONOS:TIMER_QUERY}
\section{Conclusion}
\subsection{Review}
\subsection{Future Work}
Particle Activation: \cite{MPM:GPU}

>>>>

\cite{MPM:GPU}
%look again for MPM/PIC gpu impl
%SPH, FEM methods
\section{Notation}
Multi-component types are generally printed in bold letters $\boldsymbol{A}$, $\boldsymbol{a}$. Vectors furthermore use lower-case letters $\boldsymbol{v}$, while Matrices use upper-case letters $\boldsymbol{M}$. A variable in the Lagrangian form has a subscript 0 $ _0{\boldsymbol{x}}$, the Eulerian description will be written with an 'overline' $_t\boldsymbol{x}$ if distinction is needed. Occasionally notation may be omitted where it is apparent from context.
\subsection{Einstein-Notation}
Used throughout this thesis is at instances the Einstein-Notation when using tensors and vectors. Let $\boldsymbol{a}$ and $\boldsymbol{b}$ be vectors of dimension $n$ and  $\boldsymbol{A,B,D}$ $ { m \times n }$ tensors. Repeated indices that are also defined on the variable imply component-wise operations.
$a _ { i } b _ { j }$ is multiplying component $i$ with component $j$ of vectors $\boldsymbol{a}$ and $\boldsymbol{b}$. Vector and tensor/matrix addition thus become:
\begin{equation} \label{EINSTEIN:ADD}
  c _ { i } = a _ { i } + b _ { i } \quad \text { and } \quad D _ { i j } = A _ { i j } + B _ { i j }
\end{equation}
Repeated indices that are not otherwise defined however imply summation on that index. The vector dot product becomes:
\begin{equation}
  a _ { i } b _ { i } \equiv \sum _ { i = 1 } ^ { n } a _ { i } b _ { i }
\end{equation}
Following this notation the Frobenius inner product between two second-order tensors is:
\begin{equation} \label{EINSTEIN:FROBENIUS1}
  \boldsymbol { A } : \boldsymbol { B } \equiv A _ { i j } B _ { i j } \equiv \sum _ { i = 1 } ^ { n } \sum _ { j = 1 } ^ { m } A _ { ij } B _ { ij }
\end{equation}
The Frobenius inner product between a ${r \times s \times m \times n }$ fourth-order tensor $\boldsymbol{C}$  and second-order tensor creates a second order tensor combining the ideas of \ref{EINSTEIN:ADD} and \ref{EINSTEIN:FROBENIUS1}:
\begin{equation}
  \boldsymbol { A } = A_ {ij} =  \boldsymbol {C} : \boldsymbol { B } = C _ {i j k l} B _ { k l  } = \sum _ { k = 1 } ^ { m } \sum _ { l = 1 } ^ { n } C _ { ijkl } B _ { kl }
\end{equation}
The definition of the Kronecker Delta is as follows:
\begin{equation}
  \delta _ { i j } = \left\{
    \begin{array} { l l l }
      { 1 } & {\text{if}} &  {i = j}  \\
      { 0 } & {\text{if}} &  { i \neq j }
  \end{array} \right.
\end{equation}
The Kronecker Delta as an operator is most efficiently described as a substitution:
\begin{equation}
  a _ { i } \delta _ { i j } = a _ { j }
\end{equation}
I.e. this component is only evaluated if $i = j$.
\cite{MCGINTY:CONTINUUM}
\section{Basics}
\subsection{Material Point Method}
\subsubsection{Continuum Mechanics}
Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle. Rather the continuum assumption holds: each particle represents a continuous piece of material, s.t. a microscopic view does not need to be adopted. A particular body is composed of a set of particles and can adapt different configurations due to changes in shape. These changes are cause by external or internal effects (forces etc.) on it and deform the body over time. \cite{MIT:CONTINUUM_MECHANICS} \cite{MPM:COURSE}

When modelling solids changes of quantities from one configuration (reference) $_0\boldsymbol{x}$ to another configuration (current) $_t\boldsymbol{x}$ need to be measured. In the material point method the reference configuration $^0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$). This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.
Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the initial and current configuration respectively.  Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$, which relates the initial configuration to the current configuration. If we assume for simplicity that $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations, this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
\begin{equation}
  _t\boldsymbol{x} =  {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } )
\end{equation}
If for instance the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , we can defines this mapping to be:
\begin{equation}
  _t\boldsymbol{{x}} = {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } ) = \boldsymbol{R}(t)_0\boldsymbol{x}+\boldsymbol{v}t
\end{equation}
The velocity of a material point in $_0\boldsymbol{x}$ can be defined using this mapping
\begin{equation}\label{eq:velocity}
  _0\boldsymbol{v}(_0\boldsymbol{x},t) = \frac{\partial {^t_0}\phi}{\partial t}(_0\boldsymbol{x},t)
\end{equation}
and similarly the acceleration is defined
\begin{equation}
  _0\boldsymbol{a}(_0\boldsymbol{x},t) = \frac{\partial^2 {^t_0}\phi}{\partial t^2}(_0\boldsymbol{x},t) = \frac{\partial _0\boldsymbol{v}}{\partial t}(_0\boldsymbol{x},t).
\end{equation}
It is helpful to abstract away from the reference configuration and think of it as being defined in a different fixed material space. Physically this has the impact of moving with the particle in world space. Commonly known as the Lagrangian form. It is often easier in continuum mechanics to start with a Lagrangian description and switch to a Eulerian one if needed. The Eulerian description is static: Variables of particles that are moving by are measured while staying in a fixed position.
These descriptions are different but they will yield the same measurements when related correctly. These relations for some particle quantity $f$ are called the (Lagrangian) pull back
\begin{equation}
  _0f(_0\boldsymbol{x},t)= {_tf}({^t_0}\phi(_0\boldsymbol{x},t),t)
\end{equation}
and the (Eulerian) push forward
\begin{equation}\label{eq:push_forward}
  _tf(_t\boldsymbol{x},t) =  {_0f}({^t_0}\phi^{-1}(_t\boldsymbol{x},t),t) = {_0f}({_t^0}\phi(_t\boldsymbol{x},t),t)
\end{equation}
with definitions over their respective spaces $_tf ( \cdot , t ) : \Omega ^ {  { t } } \rightarrow \mathbb { R }$, $_0f ( \cdot , t ) : \Omega ^ {   { 0 } } \rightarrow \mathbb { R }$.  To enable the operator ${^t_0}\phi$ to be homeomorphic, s.t. an inverse ${^t_0}\phi^{-1} = {_t^0}\phi$ is defined, it is assumed that no two particles will ever occupy the same space at the same time.

The difficulty in the eulerian formulation becomes apparent when differentiating (due to the chain rule):
\begin{equation}
  \frac {\partial}{\partial t} {_0f_i } ( {_0\boldsymbol { x }} ,  t  ) =
  \frac { \partial  {_tv_i} } { \partial t } ( {^t_0}\phi ( {_0\boldsymbol {x }} , t ) , t ) + \frac { \partial  {_tv_i }} { \partial {_tx_j} } ( {^t_0}\phi ( {_0\boldsymbol { x }} , t ) , t ) \frac { \partial {^t_0}\phi_j } { \partial t } ( {_0\boldsymbol { x }} , t )
\end{equation}
Combining this with equation \ref{eq:velocity} and applying the push forward \ref{eq:push_forward} to cancel out mappings leads to the definition referred to as the material derivative:
\begin{equation}
  \frac {  { D } } {   { D }   { t } }   {  {f} } ( _t\boldsymbol { x } ,   { t } ) = \frac { \partial   {  {f} } } { \partial   { t } } ( _t\boldsymbol {x } ,   { t } ) + \frac { \partial   {  {f} } } { \partial x _ {   { j } } } ( _t{\boldsymbol { x }} ,   { t } )  {v} _ {   { j } } ( _t\boldsymbol { x } ,   { t } )
\end{equation}
The Jacobian of the deformation map $\phi$ is the deformation gradient and is one of the key components to measure strain:
\begin{equation}
  {^t_0 F_{ij}} ( _0\boldsymbol {x} , \boldsymbol { t } ) = \frac { \partial {^t_0\phi_i} } { \partial _0 x _j } ( _0\boldsymbol { x } , \boldsymbol { t } ) = \frac { \partial _t x _i } { \partial _0 x_j } ( _0\boldsymbol { x } , \boldsymbol { t } )
\end{equation}
Intuitively it measures the local changes in length from the reference to the current configuration.
\cite{MIT:CONTINUUM_MECHANICS}
\cite{MPM:COURSE}
\subsubsection{Conservation Equations}
\cite{MPM:APIC}
\subsubsection{CFL Condition}
\subsection{Singular Value Decomposition}
\cite{SVD:3x3}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                         Constitutive Models                         %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Constitutive Models}
\cite{MPM:DRUCKER} %Sand Plasticity
\cite{MPM:MULTI} %Sand+Water
\cite{MPM:INVERT} %Neo-Hookean
\cite{MPM:OLROYDB} %Viscoelastic Foam,Sponge
\cite{MPM:SHELLS} %Thin Shells
\cite{MPM:PHASE_CHANGE} %Heat-Model
\cite{MPM:OPTIMI_INTEGR} %Opimization Methods, BDF-Semi-Impl-Midpoint rule

The first aim will be to find the strain energy density $\psi(\boldsymbol{\epsilon})$ of the strain $\bm{\epsilon}$.

@@

where $\boldsymbol{\sigma}$ and $\bm{\epsilon}$ are second-order tensors with (3x3) = 9 elements. $C_{ijkl}$ is a fourth-order tensor with (3x3)x(3x3) = 81 elements.

@@

  \sigma _ { i j } = \sigma _ { i j } ( \boldsymbol{\epsilon} ) = \frac { \partial { \psi } } { \partial \epsilon _ { i j } },

@@

  = \boldsymbol{C}^{-1} \bm{\sigma}

@@

using engineering constants Young's modulus $E$ and Poisson ratio $\nu$. Inverting $\boldsymbol{C}^{-1}$ and switching to Lamé parameters $\lambda$ and $\mu$ results in the equation:

@@

  \boldsymbol{\epsilon} = \frac { 1 } { 2 } \left( \boldsymbol { F }_E + \boldsymbol { F }_E ^ { T } \right) - \boldsymbol { I }

@@

The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t. to $\boldsymbol{F}_E$: $\psi(\boldsymbol{R}_{0} \boldsymbol{F}_E) \neq \psi(\boldsymbol{F}_E)$. The deformation caused by stress however does not affect rigid body motions. Given the polar decomposition $\boldsymbol{F}_E = \boldsymbol{R}_E\boldsymbol{S}_E$, an alternate strain measure may be defined as:

@@

  { \boldsymbol { \epsilon } } ( \boldsymbol { F }_E ) = \boldsymbol { \epsilon } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E \right) = \frac { 1 } { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E + \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E \right) ^ { T } \right) - \boldsymbol { I } = \boldsymbol { S }_E - \boldsymbol { I }

@@

Substituting $ {\boldsymbol{\epsilon}}$ into equation \ref{eq:linear_elas} leads to the energy definition of corotational hyperelasticity:
\begin{equation}\label{eq:corot_S}
  \Psi = \mu \| \boldsymbol{ S }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { S }_E - \boldsymbol { I } )
Using rotation-invariance of the Frobenius norm:

@@

  \Psi = \mu \| \boldsymbol {F}_E  - \boldsymbol {R}_E \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E - \boldsymbol { I } \right)

@@

Even more insight yields the relationship to their singular values $\sigma _ i$, also called the principal stretches:
$$\operatorname {tr}\left(\boldsymbol{S}\right) = \sum _ { i = 1 } \sigma _ { i } = \operatorname{tr}\left(\boldsymbol{\Sigma}_E\right)$$
$$\| \boldsymbol{S}_E \| _ { F }^2 = \sum _ { i = 1 } \sigma _ { i } ^ { 2 } = \| \boldsymbol{\Sigma}_E\|_{F}^2$$
$$\Rightarrow \| \boldsymbol{S}_E\boldsymbol{-I} \| _ { F }^2 = \|\boldsymbol{S}_E \| _ { F }^2 - 2 \operatorname{tr}(\boldsymbol{S}_E)  + \|\boldsymbol{I}\|_{F}^2 = \| \boldsymbol { \Sigma }_E - \boldsymbol { I } \| _ { F } ^ { 2 }$$
\begin{equation} \label{eq:corot_sing}
  \Psi(\boldsymbol{\Sigma}_E) = \mu \| \boldsymbol { \Sigma }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma }_E - \boldsymbol { I } )
Equation \ref{eq:corot_sing} uses the diagonal matrix $\boldsymbol{\Sigma}_E$ containing the principal stretches typically acquired by the singular value decomposition $\boldsymbol{F}_E \boldsymbol{= U \Sigma}_E\boldsymbol{ V ^T }.$ Since the energy density in \ref{eq:corot_sing} is a function of only three (singular) values, which describe stretch/compression of the material, isotropy of the material is underlined.
\cite{ADAMS:ELASTICITY}
Numerical Stepping algorithms
-> Newton Raphson scheme
-> Problem: Bad Primary contour
Benefit: Easier description in terms of Invariants
\cite{MPM:INVERT}

\subsection{GPU memory}
\subsubsection{Optimization opportunities}
\cite{NVIDIA:BEST:PRACTICE}
\cite{AMD:GPU_OPEN}
\begin{itemize}
  \item Minimize CPU-GPU transfers
  \item Coalesced memory access
  \item Bank conflicts
\end{itemize}
\subsubsection{SoA vs. AoS}
\section{Implementation}
\newcommand{\var}[1]{{\ttfamily#1}}% variable
\begin{algorithm}[t]
  \caption{Euclid’s algorithm}\label{euclid}
  \begin{algorithmic}[1]
    \Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
    \State $r\gets a\boldsymbolod b$ and \var{foobar}\label{foobar}
    \While{$r\not=0$}\Comment{We have the answer if r is 0}
    \State $a\gets b$
    \State $b\gets r$
    \State $r\gets a\boldsymbolod b$
    \EndWhile\label{euclidendwhile}
    \State \textbf{return} $b$\Comment{The gcd is b}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}
\subsection{Framework}
decorator pattern
\subsection{Parallel Reduction}
\cite{NVIDIA:PARALLEL_REDUCTION}
\subsection{Binning}
\cite{NVIDIA:BINNING}
\subsection{Scanning}
\cite{NVIDIA:SCAN}
\cite{NVIDIA:SCAN_MODERN}
\subsection{Affine particle-in-cell}
\cite{MPM:APIC}
\subsection{Iterative solvers for linear systems}
%GMRES, Conjugate Gradients
\subsection{Verifying Simulation}
\begin{itemize}
  \item Conservation equations
  \item Singular Value Decomposition
  \item Parallel Reduction
\end{itemize}
\section{Evaluation}
\cite{NVIDIA:PEAK_PERFORMANCE}
\cite{KHRONOS:TIMER_QUERY}
\section{Conclusion}
\subsection{Review}
\subsection{Future Work}
Particle Activation: \cite{MPM:GPU}
